---
title: "Tutorial: plotting maps of Spain with ggplot2"
author: "Ariane Aumaitre"
date: "27 de octubre de 2019"
output:  github_document
editor_options:
  markdown:
    wrap: 80
---

```{r global options, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE, dev = "ragg_png",
  tidy = "styler",
  dpi = 300,
  fig.path = "README-",
  out.width = "100%"
)
```

After spending the last few days playing with maps, in this blogpost I'll be
showing you how to create what I think look like nice maps using just ggplot.
While other packages may seem more map-friendly at first glance, I don't think
any of the alternatives gives the possibility of personalizing the output as
much as ggplot does - and sorry not sorry I love doing that.

Little disclaimer: a lot of the inspiration for this post comes from the amazing
[Beautiful thematic maps with
ggplot2](https://timogrossenbacher.ch/2016/12/beautiful-thematic-maps-with-ggplot2-only/)
blogpost, and I don't think my maps bring any added value to it. However, this
may still be useful if you're interested in plotting a map of *Spain*, as I'll
be discussing some specificities of that.

I'll be "drawing" two maps here: a regional and a local one. Once you get the
logic, the process is basically the same, but I think it can help in
understanding the logic incrementally. Let's go!

```{r}
library(tidyverse) # can we actually code without it?
library(sf) # to manipulate shapefiles
library(mapSpain) # to get shapes of Spain
library(wesanderson) # for some nice color palettes
```

### First things first: we need shapefiles

Maps are drawn in ggplot using **geom_sf()**. But in order to draw any of these,
we first need a `sf` object giving us the coordinates of the borders of our map.
You can find a lot of shapefiles online, as well as libraries that provide these
type of object. I'll be using for this tutorial the library `mapSpain`.

Now it's time to get the shapefiles.

```{r, message = FALSE}
# regional shapefile:
sf_regional <- esp_get_ccaa_siane()

# Lines around the Canary Islands by mapSpain
canaries_line <- esp_get_can_box()
```

Now time to repeat the same process for the local level. Here, I'll be using the
ine_cod (CODIGOINE in the database) variable to identify the municipalities,
that I need to create:

```{r}
# local shapefile:
sf_local <- esp_get_munic_siane() %>%
  mutate(ine_cod = paste0(cpro, cmun))
```

### A regional plot of intergenerational inequalities

Now that we have our shapefiles ready to be plotted, we just need some data to
represent. Here I'll be using what got me started with this in the first place:
the relative income of young people (18-25) as compared to the national average
income, by region ("Comunidad Autónoma"). This comes from the Spanish branch of
the EU SILC, that is publicly available
[here](https://www.ine.es/dyngs/INEbase/es/operacion.htm?c=Estadistica_C&cid=1254736176807&menu=ultiDatos&idp=1254735976608),
and I'm using data from 2017. That's what I was working with, but you can use
whatever you want.

**Keep in mind** that you'll need to join your data to represent with the
shapefile data, so be sure that you have a column in each of those that allows
for joining. I'm going to skip the data cleaning here and go directly to the
join, that I'll do with the id column (this was the easiest as both data frames
had the same order).

**Note that** the original `plot_data.RData` already had an `id` column
representing the numeric identifier of the "Comunidad Autónoma" but starting in
`0` instead of `1`, so I need to create a filed named `id` following that logic
in `regional_plot`.

```{r}
# importing the data
load("plot_data.RData")

# joining with the regional df
regional_plot <- sf_regional %>%
  # Logic for joining
  mutate(id = as.numeric(codauto) - 1) %>%
  left_join(plot_data, by = "id")
```

It's finally time to plot! We'll start with a very basic plot before getting
picky with the details:

```{r ugly}
regional_plot %>%
  ggplot() +
  # ri_disp is the income variable
  geom_sf(aes(fill = ri_disp), color = "white") +
  theme_minimal()
```

Let's make a short list of things we *should hate* from this map:

-   All that text in the axis. We should say goodbye to it.
-   That legend looks really... ugly, right?
-   That blues scale is not that bad, but we can probably think of something
    better.

Let's get to work and do this a bit more appealing.

#### The theme

We can move on to making things pretty. The next chunk creates a nice theme for
maps. I've done it by mixing some things I usually apply to all my plots with
some more map-specific elements taken (and slightly modified) from the Swiss map
blogpost I quoted above:

```{r}
theme_ari_maps <- function(...) {
  theme_minimal() +
    theme(
      axis.line = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      panel.grid.major = element_line(color = "#ebebe5", linewidth = 0.2),
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill = "ivory1", color = NA),
      panel.background = element_rect(fill = "ivory1", color = NA),
      legend.background = element_rect(fill = "ivory1", color = NA),
      panel.border = element_blank(),
      plot.title = element_text(size = 11, hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(size = 9, hjust = 0.5, color = "grey40"),
      plot.caption = element_text(size = 7.5, color = "grey40"),
      legend.title = element_text(color = "grey40", size = 8),
      legend.text = element_text(color = "grey40", size = 7, hjust = 0),
      legend.position.inside = c(0.7, 0.07),
      #      legend.background = element_rect(fill = alpha('white', 0.0)),
      plot.margin = unit(c(.5, .5, .2, .5), "cm"),
      panel.spacing = unit(c(2, 0.2, .2, 0.2), "cm")
    )
}
```

#### The legend

Once again, here I'm taking full inspiration from Timo Grossenbacher's work, and
this is mostly his code (again, it's
[here](https://timogrossenbacher.ch/2016/12/beautiful-thematic-maps-with-ggplot2-only/)).
The details of the legend later on in the ggplot call also come from there. Even
if my measure of relative income is continuous, I think the plot will be more
intuitive I I turn it into a "fake discrete" by dividing my data into 5
quintiles and converting it into five categories. I do this in the next chunk,
and then use the quintiles from the actual data to choose my "pretty breaks" and
create the labels. Whether you do this or not, and the breaks you choose will
fully depend on your data and personal preferences.

```{r}
# Getting the quantiles:
quantile(regional_plot$ri_disp, probs = c(.2, .4, .6, .8), na.rm = TRUE)
# This returns .76, .86, .94, 1.03

# I'm going to slightly change the breaks to make them prettier
# again, this fully depends on your preferences
pretty_breaks <- c(.75, .85, .95, 1.05)

# Getting the minimum and maximum value to surround the breaks
minVal <- min(regional_plot$ri_disp, na.rm = T)
maxVal <- max(regional_plot$ri_disp, na.rm = T)

# Puttling them together:
brks <- c(minVal, pretty_breaks, maxVal)

# Creating labels
labels <- c()
# round the extremes
for (idx in 1:length(brks)) {
  labels <- c(labels, round(brks[idx + 1], 2))
}

labels <- labels[1:length(labels) - 1]
```

Now that breaks and labels are created, the next step transforms our `ri_disp`
variable and cuts it into the categories created by the breaks. This is what
we'll be using from now on for the `fill` aes!

```{r}
regional_plot$brks <- cut(regional_plot$ri_disp,
  breaks = brks,
  include.lowest = TRUE,
  labels = labels
)

brks_scale <- levels(regional_plot$brks)
```

#### Getting there: colors

We're almost there, but first, let's choose a color palette. These days I've
been using a lot [Wes Anderson
palettes](https://github.com/karthik/wesanderson), and I think the "Zissou1",
which ranges from blue to red, is quite appropriate for this kind of incremental
variable. We have 5 categories so I'm creating a discrete scale with 5 values.

```{r}
pal <- wes_palette("Zissou1", 5, type = "discrete")
```

#### Time to plot (again)

Now that we have fixed most of our issues with the raw plot, it's time to do the
final ggplot call. I've tried to add comments to all the steps which could
create confusion.

```{r regional}
regional_plot %>%
  ggplot() +
  # we use brks for the fill and resuce the size of the borders
  geom_sf(aes(fill = brks), color = "white", linewidth = 0.3) +
  # Line to separate the Canary Islands
  geom_sf(data = canaries_line, color = "grey40") +
  # Adding the color palette
  # AND setting how I want the scale to look like
  scale_fill_manual(
    values = rev(pal), # I use rev so that red is for lowest values
    # breaks = brks_scale,
    name = "Renta relativa",
    drop = FALSE,
    guide = guide_legend(
      direction = "horizontal",
      position = "inside",
      nrow = 1,
    )
  ) +
  labs(
    title = "La brecha territorial generacional, por territorios",
    subtitle = "Nivel de renta de los jóvenes (18-25) relativo a la media de la población en España, 2017",
    caption = "Ariane Aumaitre - Datos: ECV"
  ) +
  theme_ari_maps() +
  # ggplot2 v3.5.0 for legend
  theme(
    legend.position = "bottom",
    legend.title.position = "top",
    legend.title = element_text(hjust = 0),
    legend.text.position = "bottom",
    legend.text = element_text(hjust = 1),
    legend.key.width = unit(50 / length(labels), units = "mm"),
    legend.key.height = unit(2, units = "mm")
  )
```

## Average income at the local level

Last but not least, we'll be applying all the logic from above to create a map
at the local level. I'll be using data on **average income at the local level**
(variable: `renta`) that can be downloaded from
[here](https://www.ine.es/experimental/atlas/exp_atlas_tab.htm). The steps to be
followed are the same:

-   Importing the data (I'm skipping the data cleaning again)
-   Joining it with the shapefile data frame (by variable `ine_cod`)
-   Bring the Canary Islands closer
-   Playing around a bit to get nice legend breaks
-   Get our theme and plot!

```{r}
# Loading data
load("income_data.RData")

# Joining and editing the Canary Islands position
local_plot <- sf_local %>%
  left_join(income_data, by = "ine_cod") %>%
  mutate(
    renta = ifelse(is.na(renta), NA, as.numeric(renta)) # was imported as character
  )


# Getting quintiles to decide on breaks
quantile(local_plot$renta, probs = c(.2, .4, .6, .8), na.rm = TRUE)

# This returns 8309, 9538, 10674, 12083


# Now I repeat the whoe exact break and labels process again
# See regional plot for explanation
pretty_breaks <- c(8500, 9500, 10500, 12000)
minVal <- min(local_plot$renta, na.rm = T)
maxVal <- max(local_plot$renta, na.rm = T)



# All together:
brks <- c(minVal, pretty_breaks, maxVal)

# Labels
labels <- c()
# round the extremes
for (idx in 1:length(brks)) {
  labels <- c(labels, round(brks[idx + 1], 2))
}

labels <- labels[1:length(labels) - 1]

# Bringing into df
local_plot$brks <- cut(local_plot$renta,
  breaks = brks,
  include.lowest = TRUE,
  labels = labels
)

brks_scale <- levels(local_plot$brks)
```

And time to create the final plot!!

```{r local}
local_plot %>%
  ggplot() +
  # No borders this time
  geom_sf(aes(fill = brks), color = "white", linewidth = 0.1) +
  # Line to separate the Canary Islands
  geom_sf(data = canaries_line, color = "grey40", alpha = 0.7) +
  # Adding the color palette
  # AND setting how I want the scale to look like
  scale_fill_manual(
    values = rev(pal), # I use rev so that red is for lowest values
    breaks = brks_scale,
    name = "Renta media (€)",
    drop = FALSE,
    guide = guide_legend(
      direction = "horizontal",
      position = "inside",
      nrow = 1,
    )
  ) +
  labs(
    title = "La brecha territorial en España",
    subtitle = "Renta relativa por municipio con respecto a la renta media nacional, 2016",
    caption = "Ariane Aumaitre - Datos: INE"
  ) +
  theme_ari_maps() +
  # ggplot2 v3.5.0 for legend
  theme(
    legend.position = "bottom",
    legend.title.position = "top",
    legend.title = element_text(hjust = 0),
    legend.text.position = "bottom",
    legend.text = element_text(hjust = 1),
    legend.key.width = unit(50 / length(labels), units = "mm"),
    legend.key.height = unit(2, units = "mm")
  )
```
